Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase Fix Time?

### 2015-03-17 data.tex
Updates from ShareLaTeX

\subsection{Overview of the Team Software Process}
To empirically test this heuristic, we examined the phase delay effect in software projects conducted between 2005 and 2014. These projects took place at organizations in many countries and were conducted using  the Team Software Process (TSP$\textsuperscript{SM}$ )), a software project management approach developed at the Software Engineering Institute (SEI) at Carnegie Mellon University. TSP is an extension of the Personal Software Process (PSP$\textsuperscript{SM}$ )) developed at the SEI by Watts Humphrey ~\cite{tsp00}. The data from these TSP projects were collected and stored in the Software Engineering Measured Process Repository (SEMPR) at the SEI. The Software Engineering Institute (SEI) at Carnegie Mellon University explores methods for software process improvement.

One product from that research is  
Watts Humphrey's  Team Software Process$\textsuperscript{SM}$  (TSP$\textsuperscript{SM}$ )~\cite{tsp00} which extends Humphrey's early
work on Personal Software
Process$\textsuperscript{SM}$  (PSP$\textsuperscript{SM}$ )~\cite{psp05}.

TSP has guiding principles; the following are relevant to this study:
\begin{itemize}
\item  Sound engineering work does not happen by accident. It must be planned.
\item  Performance of the plans requires realistic commitments of the resources needed to perform the work.
\item  The planned work should be compared to actual frequently to assure plans remain relevant.
\item  People performing the work should be responsible for planning and tracking the work.
\end{itemize}

@@

Common features of TSP projects include planning, personal reviews, peer inspections, and coaching.

A TSP coach helps the team to plan and analyze performance. The coach is the only role authorized to submit project data to the SEI.
Before reviewing data with the teams, therefore before submission, these coaches check the data for obvious errors.

Planning follows the technique taught in the PSP$\textsuperscript{SM}$. Developers estimate the size of work products and convert this to a total effort using historical rates. Time in specific tasks come from the  process phases and historical percent time in phase distributions. Defects are estimated using historical phase injection rates and phase removal yields. Coaches help the developers to compare estimates against actual results. In this way, developers acquire a more realistic understanding of their work behavior, performance, and schedule status.

Personal review is a technique taken from the PSP and its use in TSP is unique.  Developers follow a systematic process to remove defects by  examining their own work products using a checklist built from their personal defect profile. This personal review occurs after some product or part of a product is considered to be constructed and before peer reviews or test.

@@

Peer inspection is a  technique in
traditional software engineering and is often called peer review.
 Basili and Boehm   commented in 2001~\cite{boehm01} 
that peer reviews can catch over half the defects introduced into a system.
Peer inspection can be conducted on any artifact generated anywhere in the software
lifecycle and can quickly be adapted to new kinds of artifacts. TSP peer reviews follow the Fagan style in which the reviewer uses a checklist composed of common team defects prior to a review team meeting.

\subsection{Projects in the Sample}
Since shortly before the year 2000, the SEI has been teaching and coaching TSP teams. One of the authors (Nichols) has been been mentoring software development teams and coaches around the world as they deploy TSP within their organizations since 2006.  The earliest projects in our sample began work in 2004; the most recent completions were in 2014.
The projects were mostly small to medium, with a median duration of 46 days and a maximum duration of 90 days in major increments. 
Several projects extended for multiple incremental development cycles. 
Median team size was 7 people, with a maximum of 40.

Many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
There were  some  medical device projects in  the US, France, Japan, and Germany as well  commercial 
computer-aided design systems, and embedded systems.

An anonymized version of that data is available in the PROMISE repository  and,
for confidentiality restrictions, we cannot offer 
further details on these projects.

\tim{Needs either a URL or a reference.}

@@

To use this data for research and benchmarking,  Nichols first collected the Process Dashboard data from the submission repository then used a tool built by the Process Dashboard developer, David Tuma, 
to gather the project data into a database. Nichols and his colleague Yasutaka Shirai then extracted data into views suitable for analysis. The key views include a project results summary, effort logs, task logs, and defect logs. The data has not yet undergone additional screening. A summary of the data quality issues was reported in a previous work ~\cite{shirai14}.

@@

Note that, in that Figure~\ref{fig:waterfall}:
\bi 
\item
Several  phases in which product is created have sub-phases of {\em review} and {\em inspect} to remove defects. TSP uses review, a sub-phase in which individuals perform personal reviews of their work products prior to the peer review, which TSP calls the inspection.
\item Testing is divided into several stages. Developers perform unit test prior to code complete.  After code complete a standard phase is the integration, which combines program units into workable system ready for system test. Integration,  system test, and acceptance test are often performed by a separate group. 
\ei

The principles followed are that 
\bi 
\item Defects should be found before test, that is, test defects should be infrequent.
\item A product should be inspected before used in a subsequent construction phase to minimize  unnecessary rework.
\item To minimize unnecessary rework, a product should be reviewed and inspected prior to use in a subsequent construction phase.
\item A personal review should remove the most common defects before being given to peers for the inspection.
\item Because different categories of test will find different types of defect, and follow in a natural sequence, the tests should be measured separately
\ei

In summary, the TSP process and measurement have been specifically designed to aid the analysis of defects for use in process improvement. The data are, therefore, uniquely suitable for an analysis of the phase delay effect. 
\subsection{Distinguishing Project Characteristics}
Modern development processes include the core engineering practices of estimation, planning, configuration control, requirements, design, code development, and test. Agile methods address  not only these core practices but also address team organization, individual and team commitments, test and build automation, and implementing iterative and incremental development. TSP includes using additional practices associated with high performance ~\cite{jones10} including; project tracking and control, specialization of team members, inspections, static analysis, documentation and training, and reuse. Although the teams in this sample use many typical agile practices, they are also using additional practices.

>>>>

\subsection{Overview of the Team Software Process}
To empirically test for the phase delay effect, we examined 171 software projects conducted between 2005 and 2014. These projects took place at organizations in many countries and were conducted using  the Team Software Process (TSP$\textsuperscript{SM}$), a software project management approach developed at the Software Engineering Institute (SEI) at Carnegie Mellon University~\cite{tsp00}. TSP is an extension of the Personal Software Process (PSP$\textsuperscript{SM}$) developed at the SEI by Watts Humphrey ~\cite{tsp00}. The data from these TSP projects were collected and stored in the Software Engineering Measured Process Repository (SEMPR) at the SEI. The Software Engineering Institute (SEI) at Carnegie Mellon University explores methods for software process improvement.

TSP has guiding principles; the following are relevant to this study:
\begin{itemize}
\item  Sound engineering work does not happen by accident. It must be planned.
\item  Performance of the plans requires realistic commitments of the resources needed to perform the work.
\item  The planned work should frequently be compared to  results to assure plans remain relevant.
\item  People performing the work should be responsible for planning and tracking the work.
\end{itemize}

@@

Common features of TSP projects include {\em planning}, {\em personal reviews}, {\em peer inspections}, and {\em coaching}.
A TSP {\em coach} helps the team to plan and analyze performance. The coach is the only role authorized to submit project data to the SEI.
Before reviewing data with the teams, therefore before submission, these coaches check the data for obvious errors.

{\em Planning} follows the technique taught in the PSP$\textsuperscript{SM}$. Developers estimate the size of work products and convert this to a total effort using historical rates. Time in specific tasks come from the  process phases and historical percent time in phase distributions. Defects are estimated using historical phase injection rates and phase removal yields. Coaches help the developers to compare estimates against actual results. In this way, developers acquire a more realistic understanding of their work behavior, performance, and schedule status.

{\em Personal review} is a technique taken from the PSP and its use in TSP is unique.  Developers follow a systematic process to remove defects by  examining their own work products using a checklist built from their personal defect profile. This personal review occurs after some product or part of a product is considered to be constructed and before peer reviews or test.

@@

{\em Peer inspection} is a  technique in
traditional software engineering and is often called peer review.
 Basili and Boehm   commented in 2001~\cite{boehm01} 
that peer reviews can catch over half the defects introduced into a system.
Peer inspection can be conducted on any artifact generated anywhere in the software
lifecycle and can quickly be adapted to new kinds of artifacts. TSP peer reviews follow the Fagan style in which the reviewer uses a checklist composed of common team defects prior to a review team meeting.

\begin{figure*}[!t]
\begin{center}
\includegraphics[width=6in]{waterfall-v2.png}
\end{center}
\caption{Different approaches to software development:  waterfall and agile (bottom left).}
\label{fig:waterfall}
\end{figure*}

\subsection{Projects in the Sample}
Since shortly before the year 2000, the SEI has been teaching and coaching TSP teams. One of the authors (Nichols) has been been mentoring software development teams and coaches around the world as they deploy TSP within their organizations since 2006.  The  most recent completions were in 2014.
The projects were mostly small to medium, with a median duration of 46 days and a maximum duration of 90 days in major increments. 
Several projects extended for multiple incremental development cycles. 
Median team size was 7 people, with a maximum of 40.

Many of the projects were e-commerce web portals or banking systems in the US, South Africa, and Mexico. 
There were  some  medical device projects in  the US, France, Japan, and Germany as well  as a commercial computer-aided design systems, and embedded systems.

An anonymized version of that data is available in the PROMISE repository\footnote{http://openscience.us/repo}  and,
for confidentiality restrictions, we cannot offer 
further details on these projects.

@@

To use this data for research and benchmarking,  Nichols first collected the Process Dashboard data from the submission repository then used a tool built by the Process Dashboard developer, David Tuma, 
to gather the project data into a database. Nichols and his colleague Yasutaka Shirai then extracted data into views suitable for analysis. The key views include a project results summary, effort logs, task logs, and defect logs. The data has not yet undergone additional screening. A summary of the data quality issues was reported in a previous work ~\cite{shirai14} (that summary is discussed further in our {\em Validity} section, below).

@@

Note that, in that Figure~\ref{fig:waterfall}:
\bi 
\item
Several  phases in which product is created have sub-phases of {\em review} and {\em inspect} to remove defects. TSP uses review, a sub-phase in which individuals perform personal reviews of their work products prior to the peer review, which TSP calls the inspection.
\item Testing is divided into several stages. Developers perform unit test prior to code complete.  After code complete a standard phase is the integration, which combines program units into workable system ready for system test. Integration,  system test, and acceptance test are often performed by a separate group. 
\ei
For these TSP projects, the principles followed are: 
\bi 
%\item Defects should be found before test, that is, test defects should be infrequent.
\item A product should be inspected before used in a subsequent construction phase to minimize  unnecessary rework.
\item To minimize unnecessary rework, a product should be reviewed and inspected prior to use in a subsequent construction phase.
\item A personal review should remove the most common defects before being given to peers for the inspection.
\item Because different categories of test will find different types of defect, and follow in a natural sequence, the tests should be measured separately
\ei 
In summary, the TSP process and measurement have been specifically designed to aid the analysis of defects for use in process improvement. The data are, therefore, uniquely suitable for an analysis of the phase delay effect. 
\subsection{Distinguishing Project Characteristics}
Modern development processes include the core engineering practices of estimation, planning, configuration control, requirements, design, code development, and test. Agile methods address  not only these core practices but also address team organization, individual and team commitments, test and build automation, and implementing iterative and incremental development. TSP uses additional practices associated with high performance ~\cite{jones10} including; project tracking and control, specialization of team members, inspections, static analysis, documentation and training, and reuse. Although the teams in this sample use many typical agile practices, they are also using additional practices.

### 2015-03-17 lucas.tex
Updates from ShareLaTeX

We conducted a survey of software engineering practitioners and researchers on their views of commonly-held software engineering ``laws''. One of the laws questioned relates directly to phase delay: ``Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}). We abbreviate this law as RqtsErr. The survey was conducted in two phases using Amazon's Mechanical Turk. The first phase was conducted only with professional software engineers, and the second survey was conducted only with Program Committee (PC) members of the ESEC/FSE and ICSE conferences.

@@

\begin{figure}[!ht]
\renewcommand{\baselinestretch}{0.7}
\scriptsize
\begin{center}
\subcaption{Practitioner survey}
\begin{tabular}{l|c|c|c|c|c}
Law & N & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\ 
 & & $\bar{x}$ & $\tilde{x}$ & $\bar{x}$ & $\tilde{x}$ \\
\hline 
\textbf{Rqts errors are most expensive...} & 16 & 4.6 & 5 & 4.2 & 4 \\ 
Process maturity improves output & 17 & 4.4 & 4 & 3.9 & 4 \\ 
Most time is spent removing errors & 16 & 4.1 & 4 & 3.9 & 4 \\ 
Missing reqts are hardest to fix & 17 & 4.1 & 4 & 4.0 & 4 \\
Reuse increases prod. and qual. & 16 & 3.9 & 4 & 3.8 & 4 \\
OO-programming reduces errors & 13 & 3.9 & 4 & 3.5 & 4 \\
80-20 rule (defects to modules) & 12 & 3.8 & 4 & 4.0 & 4 \\
Adding manpower to a late project & 15 & 3.7 & 4 & 3.7 & 4 \\
Inspections can remove 90\% of defects & 18 & 3.3 & 4 & 3.5 & 4 \\
Smaller changes have higher error density & 14 & 2.8 & 3 & 3.4 & 3.5 \\
A developer is unsuited to test own code & 17 & 2.6 & 3 & 3.5 & 4
\end{tabular} 
\bigskip
\subcaption{Researcher survey}
\begin{tabular}{l|c|c|c|c|c}
Law & N & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\ 
 & & $\bar{x}$ & $\tilde{x}$ & $\bar{x}$ & $\tilde{x}$ \\
\hline 
Process maturity improves output & 4 & 4.0 & 4 & 4.0 & 4 \\
\textbf{Rqts errors are most expensive...} & 30 & 3.8 & 4 & 3.8 & 4   \\ 
80-20 rule (defects to modules) & 6 & 3.7 & 4 & 3.5 & 4 \\
\textbf{PhaseDelay} & 30 & 3.6 & 4 & -- & --  \\ 
Missing reqts are hardest to fix & 7 & 3.6 & 4 & 3.7 & 4 \\
Adding manpower to a late project & 4 & 3.5 & 4 & 3.5 & 4 \\
Inspections can remove 90\% of defects & 7 & 3.4 & 4 & 3.6 & 3 \\
Reuse increases prod. and qual. & 6 & 3.3 & 4 & 3.0 & 4 \\
OO-programming reduces errors & 6 & 3.2 & 4 & 2.7 & 3 \\
Most time is spent removing errors & 6 & 2.8 & 3 & 3.2 & 4 \\ 
Smaller changes have higher error density & 4 & 2.8 & 3 & 4.0 & 4 \\
A developer is unsuited to test own code & 7 & 2.1 & 2 & 2.4 & 3
\end{tabular}

>>>>

Our survey collected data on software engineers' views of commonly held software engineering ``laws''.
One of the laws questioned relates directly to phase delay: ``Requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' (from Glass~\cite{glass02} p.71 who references Boehm \& Basili~\cite{boehm01}). We abbreviate this law as RqtsErr.

The survey was conducted in two phases using Amazon's Mechanical Turk. The first phase was conducted only with professional software engineers solicited through the Mechanical Turk\footnote{Professional software engineers were required to complete a pretest to verify their status as a professional or open source software developer and to confirm their knowledge of basic software engineering terminology and technology.}, and the second survey was conducted Program Committee members of the ESEC/FSE and ICSE conferences solicited via email.

@@

\begin{figure}[!ht] 
\scriptsize
\begin{center}
\subcaption{Practitioner survey}
\begin{tabular}{l|c|c|c|c|c}
Law & N & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\ 
 & & $\bar{x}$ & $\tilde{x}$ & $\bar{x}$ & $\tilde{x}$ \\
\hline 
\textbf{Rqts errors are most expensive...} & 16 & 4.6 & 5 & 4.2 & 4 \\ 
Process maturity improves output & 17 & 4.4 & 4 & 3.9 & 4 \\ 
Most time is spent removing errors & 16 & 4.1 & 4 & 3.9 & 4 \\ 
Missing reqts are hardest to fix & 17 & 4.1 & 4 & 4.0 & 4 \\
Reuse increases prod. and qual. & 16 & 3.9 & 4 & 3.8 & 4 \\
OO-programming reduces errors & 13 & 3.9 & 4 & 3.5 & 4 \\
80-20 rule (defects to modules) & 12 & 3.8 & 4 & 4.0 & 4 \\
Adding manpower to a late project & 15 & 3.7 & 4 & 3.7 & 4 \\
Inspections can remove 90\% of defects & 18 & 3.3 & 4 & 3.5 & 4 \\
Smaller changes have higher error density & 14 & 2.8 & 3 & 3.4 & 3.5 \\
A developer is unsuited to test own code & 17 & 2.6 & 3 & 3.5 & 4
\end{tabular} 
\bigskip
\subcaption{Researcher survey}
\begin{tabular}{l|c|c|c|c|c}
Law & N & \multicolumn{2}{c}{agreement} & \multicolumn{2}{c}{applicability} \\ 
 & & $\bar{x}$ & $\tilde{x}$ & $\bar{x}$ & $\tilde{x}$ \\
\hline 
Process maturity improves output & 4 & 4.0 & 4 & 4.0 & 4 \\
\textbf{Rqts errors are most expensive...} & 30 & 3.8 & 4 & 3.8 & 4   \\ 
80-20 rule (defects to modules) & 6 & 3.7 & 4 & 3.5 & 4 \\
\textbf{PhaseDelay} & 30 & 3.6 & 4 & -- & --  \\ 
Missing reqts are hardest to fix & 7 & 3.6 & 4 & 3.7 & 4 \\
Adding manpower to a late project & 4 & 3.5 & 4 & 3.5 & 4 \\
Inspections can remove 90\% of defects & 7 & 3.4 & 4 & 3.6 & 3 \\
Reuse increases prod. and qual. & 6 & 3.3 & 4 & 3.0 & 4 \\
OO-programming reduces errors & 6 & 3.2 & 4 & 2.7 & 3 \\
Most time is spent removing errors & 6 & 2.8 & 3 & 3.2 & 4 \\ 
Smaller changes have higher error density & 4 & 2.8 & 3 & 4.0 & 4 \\
A developer is unsuited to test own code & 7 & 2.1 & 2 & 2.4 & 3
\end{tabular}

### 2015-03-17 lucasRelatedWork.tex
Updates from ShareLaTeX

The first data on the cost to fix defects as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}). These studies suggest that the cost (in terms of effort) to find and fix an error monotonically increases with lifecycle phase. For large projects, the ratio of cost-to-fix in the requirements phase to cost-to-fix in operation is on the order of 1:100. Furthermore, cost function is superlinear, with the greatest rates of increase in the acceptance testing and operations phases.

@@

In the 40 years since these initial studies, few studies have been published on the cost-to-fix curve as a function of lifecycle phase. Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects (from two student projects of 2000 deliverable source instructions) is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}). Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects. Workshop participants from Toshiba and IBM reported cost-to-fix ratios of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- the raw data points are not provided. One notable example to the traditional cost-to-fix curve is the CCPDS-R described by Royce~\cite{Royce98}: a million-line, safety-critical missile defense system (Figure~\ref{fig:royce}). In this project, design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly. Boehm~\cite{Boehm10} attributes this success to the CCPDS-R development process, which focused on removing architecture risk early in the development lifecycle.

\begin{figure}[!ht]
 \includegraphics[width=3.3in]{Royce98.png}
 \caption{Exception to the rule - CCPDS-R case study cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
 \end{figure}

@@

\begin{figure}[!ht]
 \includegraphics[width=3.3in]{clutterbuck.png}
 \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
 \end{figure}

>>>>

One reason that industrial practitioners and academics believe so strongly in the phase delay effect,
and the impacts of requirement errors, is that they are often presented in the SE literature. For example,
we note that the gradaute SE class at North Carolina State University has presented to students at
least one  slide on the phase delay
effect (without quarrel or critical comment), every year for the last decade.

Yet when we look at the literature, the evidence for phase delay is both very sparse and very old.
The first data on the cost to fix defects as a function of lifecycle phase date back to large systems in the late 70s from IBM~\cite{Fagan76}, TRW~\cite{Boehm76}, GTE~\cite{Daly77}, and Bell Labs~\cite{Stephenson76} (Figure~\ref{fig:cost-to-fix}). These studies suggest that the cost (in terms of effort) to find and fix an error monotonically increases with lifecycle phase. For large projects, the ratio of cost-to-fix in the requirements phase to cost-to-fix in operation is on the order of 1:100. Furthermore, cost function is superlinear, with the greatest rates of increase in the acceptance testing and operations phases.

@@

In the 40 years since these initial studies, few studies have been published on the cost-to-fix curve as a function of lifecycle phase. Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects (from two student projects of 2000 deliverable source instructions) is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}). Shull et al.~\cite{Shull02} conducted a literature survey and held a series of e-workshops with industry experts on fighting defects. Workshop participants from Toshiba and IBM reported cost-to-fix ratios between early lifecycle and post-delivery defects of 1:137 and 1:117 for large projects respectively~\cite{Shull02} -- the raw data points are not provided. One notable example to the traditional cost-to-fix curve is the CCPDS-R described by Royce~\cite{Royce98}: a million-line, safety-critical missile defense system (Figure~\ref{fig:royce}). In this project, design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly. Boehm~\cite{Boehm10} attributes this success to the CCPDS-R development process, which focused on removing architecture risk early in the development lifecycle.

\begin{figure}[!t]
 \includegraphics[width=3.3in]{Royce98.png}
 \caption{Exception to the rule - CCPDS-R case study cost-to-fix curve. From~\cite{Royce98}.}\label{fig:royce}
 \end{figure}

\begin{figure}[!t]
 \includegraphics[width=3.3in]{clutterbuck.png}
 \caption{Cost of change from an agile case study. From~\cite{Clutterbuck09}.}\label{fig:clutterbuck}
 \end{figure}

@@
